---
title: ''
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### MAT 115
\vspace*{-0.1in}
### DSD #2

With your randomly assigned partner, debug (fix) this code. In your *own words* explain the errors. Submit ONE answer per group as a pdf.

```{r}
require(dslabs)

# subset heights dataset to include ONLY females
#heights_f <- subset(heights, sex="Female") should be
heights_f <- subset(heights, sex=="Female")
heights_f

# make a histogram of the height variable in your newly subsetted data frame
#hist(height) #should be
hist(heights_f$height)
```

A: On line 23 it was originally heights_f <- subset(heights, sex="Female") and it should be heights_f <- subset(heights, sex=="Female"). The error was the = sign. It should be == because that is the comparison operator. The regular = sign is the assignment operator.

The second error was on line 26. It was originally hist(heights). it should be hist(heights_f$heights) because it first needs to be using the subset and second, it needs to actually access the height variable.
